---
title: "VIVC database mining"
author: "Pietro Delfino"
date: "2018-03-28"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
I downloaded manually (argh!) all the files you can find in the **data** folder (in this GitHub repo), from the website <http://www.vivc.de>, specifically all the SSR (microsatellite markers) info available and the passport data (that is, color of the berry and country of origin of the variety) for all the cultivars

If someone out there would be so kind to help me in fetching data on this website in an automated and recursive way, with any of the wonderful **R** or **Python** libraries available, it would be of great help.
The problem is that to download data from vivc.de one needs to generate an *Excel* table selecting a certain number of varieties and it is not possible to get all the info at once

Apart from that, I did the job for you (it did not take too long actually =) ) so here are the data

### Data preparation

First load some libraries

```{r, message=FALSE}
library(prettydoc); library(kableExtra); library(data.table); library(knitr); library(xlsx); library(readxl); library(dplyr); library(ggplot2); library(poppr); library(ggsci); library(forcats); library(scales)
```


```{r}
length(dir('data/DB_complete/', recursive = T))
length(dir('data/SSR_data/', recursive = T))
head(dir('data/DB_complete/', recursive = T))
head(dir('data/SSR_data/', recursive = T))
```

Let's load the complete database, listing all the files with '.xlsx' in the *DB_complete* subfolder and applying the `read.xlsx` function to this list

```{r}
vivc.list <- list.files(path = 'data/DB_complete/',
                        pattern = '.xlsx',
                        full.names = T,
                        recursive = T)

length(vivc.list)

vivc.files <- lapply(vivc.list, function(x) read.xlsx(x, sheetName  = 'Worksheet'))

names(vivc.files) <- vivc.list

length(vivc.files)
```

Let's check the structure of a single file

```{r, echo=F}
kable(head(vivc.files[[1]]), align = 'l', row.names = F)

kable(head(vivc.files[[129]]), align = 'l', row.names = F)
```

Since all the files have the same structure we can call `rbind` on the list

```{r}
vivc.all <- do.call('rbind', vivc.files)

kable(head(vivc.all), align = 'l', row.names = F)

dim(vivc.all)

rownames(vivc.all) <- c(1:12862)

DT::datatable(vivc.all, rownames = F, filter = "top")
```

Check if there are duplicated entries

```{r}
dim(vivc.all[duplicated(vivc.all), ])
```

There are no duplicated entries.
Let's plot some statistics about berry color

```{r, dev='svg'}
ggplot(vivc.all, aes(Color.of.berry.skin)) +
  geom_bar() +
  labs(x="Color of berry skin", y="Count") +
  theme_bw()
```

And some about country of origin

```{r, dev='svg'}
ggplot(vivc.all, aes(fct_infreq(Country.of.origin.of.the.variety))) +
  geom_bar() +
  labs(x="Country of origin", y="Count") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=90,hjust=1, size = 4))
```

**Italy rules over France!!**

At this stage I want to load the SSR info and merge them with the entire database. Same steps as before

```{r}
ssr.list <- list.files(path = 'data/SSR_data/',
                       pattern = '.xlsx',
                       full.names = T)

ssr.files <- lapply(ssr.list, function(x) read.xlsx(x, sheetName  = 'Worksheet'))

names(ssr.files) <- ssr.list

length(ssr.files)
```

Let's check the structure of a single file

```{r}
kable(ssr.files[[1]][1:6,1:8], align = 'l', row.names = F)

kable(ssr.files[[206]][1:6,1:8], align = 'l', row.names = F)

# same structure over all files, call rbind

ssr.all <- do.call('rbind', ssr.files)

kable(ssr.all[1:6,1:8], align = 'l', row.names = F)

dim(ssr.all)

str(ssr.all)
```

There's a problem with the column VVMD32A2, a single value is saved as two allele size. Since I don't know which one is correct, I replace it with missing value

```{r, message=FALSE}
ssr.all$VVMD32A2 <- gsub('272/274', 'NA', ssr.all$VVMD32A2)

ssr.all$VVMD32A2 <- as.numeric(ssr.all$VVMD32A2)

rownames(ssr.all) <- c(1:3688)

DT::datatable(ssr.all, rownames = F, filter = 'top')
```

Check if there are duplicated cultivars

```{r}
dim(ssr.all[duplicated(ssr.all), ])

sum(duplicated(ssr.all$Reference.variety))

#kable(head(ssr.all[duplicated(ssr.all[,1]),]), align='l', row.names = F)

#ssr.all[duplicated(ssr.all$Reference.variety),]
```

There are some duplicated names. I want to keep all of them and give different names to the ones with same name. Later I will explore if same name means also same SSR profile

```{r}
rownames(ssr.all) = make.names(names = ssr.all$Reference.variety, unique=TRUE)

#ssr.all[duplicated(ssr.all$Reference.variety),] %>%
#  .[order(as.character(.$Reference.variety)),] %>%
#  kable(., align='l', row.names = T)
```

Now the duplicated have different names, and I replace the old names with the new ones with different numbers corresponding to more synonyms

```{r, echo=F}
ssr.all$Reference.variety <- rownames(ssr.all)

kable(ssr.all[1:6,1:8], align = 'l', row.names = F)
```

Now I can merge the complete db with the SSR file. Before I check if there are duplicated names and change the names of the complete db with the same method as before, so to uniform it with the SSR file

```{r}
sum(duplicated(vivc.all$Prime.name))
```

There are 86 entries with the same name. I create unique names adding a number to the duplicated entries

```{r}
rownames(vivc.all) = make.names(names = vivc.all$Prime.name, unique=TRUE)

vivc.all$Prime.name <- rownames(vivc.all)

kable(vivc.all[1:6,], align = 'l', row.names = F)

sum(duplicated(vivc.all$Prime.name))
```

Now I can merge the two files, keeping only the entries in common

```{r}
ssr.with.info <- merge(ssr.all, vivc.all, by.x='Reference.variety', by.y='Prime.name', all=F)

dim(ssr.with.info)
```

In this way we exclude around 400 cultivars with no complete info. Let's double check also for duplicated entries

```{r}
kable(ssr.with.info[1:6,c(1:2,19:22)], align = 'l', row.names = F)

sum(duplicated(ssr.with.info))

sum(duplicated(ssr.with.info$Reference.variety))
```

### First steps with *adegenet* functions within the *poppr* package ('poppr' in CRAN)

Before converting the data frame to a **genind** object that will be used in *adegenet*, I want to check how many unique country of origins exist, since they will be used as populations for the next analysis. The idea is to group different countries within the same 'population'. To this aim I manually checked all the countries and tried to identify major populations

```{r}
kable(table(ssr.with.info$Country.of.origin.of.the.variety), align = 'l', row.names = F)

length(table(ssr.with.info$Country.of.origin.of.the.variety))
```

The number of distinct countries is 71.
Outside `R`, after long hours of meditation, I created a new country coding scheme, that is found in the *data* folder

```{r}
new.country.codes <- read.table('data/new_countries_code.txt',
                                col.names = c('old', 'new'),
                                header = F,
                                sep='\t')

kable(new.country.codes, align = 'l', row.names = F)

length(table(new.country.codes$new))
```

There are now 13 unique 'populations'. If you want to know what 'STAN' stands for, check this code conversion table

Table Header  | Second Header
------------- | -------------
CE            | Center Europe
EE            | Eastern Europe
NE            | North Europe
NoA           | North Africa
SA            | South Africa
SE            | Southern Europe
STAN          | All the countries of middle-east which ends in 'stan' (except AZERBEIJAN)
SUSA          | South and Center America
WE            | Western Europe

Now we merge the new code file with the SSR data frame to have the new 13 country codes all together with the cultivars info

```{r}
ssr.complete <- merge(ssr.with.info, new.country.codes,
                      by.x = 'Country.of.origin.of.the.variety',
                      by.y = 'old',
                      all = T)

kable(ssr.complete[1:6, c(1:4,23)], align = 'l', row.names = F)

dim(ssr.complete)
```

Using the option `all = T` with the `merge` command created some entries with `NA` in the names. Let's remove them

```{r}
ssr.complete <- ssr.complete[!is.na(ssr.complete$Reference.variety), ]

sum(is.na(ssr.complete$Reference.variety))

kable(ssr.complete[1:6, c(1:4,23)], align = 'l', row.names = F)
```

Since the data frame is formatted with two columns for each locus (one allele each column), we first need to convert it to the format: one column each locus, with the two alleles separated by a slash (/). Then we will use the function `df2genind` from *adegenet* to convert the df to a genind object

There are 9 SSR in the database; since grapevine is diploid each SSR locus has two alleles, so 18 columns in total. We create 9 new columns where we paste the content of each adjacent columns (being the df organized in this way, that is the two alleles of the same locus in adjacent columns)

```{r}
col.to.add <- 24:32

ssr.complete[, col.to.add] <- (sapply(seq(from = 3,
                                          to = 20,
                                          by=2),
                                      function(i) paste0(ssr.complete[,i],'/',
                                                         ssr.complete[,i+1]))
                               )

ssr.df.for.adegenet <- ssr.complete[,c(2,23:32)]

names(ssr.df.for.adegenet)[2] <- 'pop'

names(ssr.df.for.adegenet)[3:11] <- c('VVSA2', 'VVMD5', 'VVMD7', 'VVMD25', 'VVMD27', 'VVMD28', 'VVMD32', 'VrZAG62', 'VrZAG79') 
```

Now the new df is ready to be converted to a genind object

```{r}
ssr.complete.genind <- df2genind(ssr.df.for.adegenet[,3:11],
                                 ploidy=2,
                                 sep='/', 
                                 pop = ssr.df.for.adegenet$pop, 
                                 ind.names = ssr.df.for.adegenet$Reference.variety, 
                                 loc.names = names(ssr.df.for.adegenet)[3:11]
                                 )

ssr.complete.genind

length(unique(pop(ssr.complete.genind)))

gac <- genotype_curve(ssr.complete.genind, sample = 1000, quiet = TRUE)
```















